---
title: "NMT-Seq: Association between accessibility conservation and expression"
output: 
  BiocStyle::html_document: 
    fig_width: 12
    fig_height: 8
---

```{r echo=FALSE, include=FALSE}
suppressPackageStartupMessages(library(BPRMeth))
suppressPackageStartupMessages(library(scater))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(weights))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(cowplot))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(scales))
suppressPackageStartupMessages(library(grid))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(truncnorm))
```

```{r echo=FALSE, include=FALSE}
# Define ggplot2 theme for scatter plots
scatter_theme <- function(){
  p <- theme(
      plot.title=element_text(size=28, face='bold', margin=margin(0,0,10,0), hjust=0.5),
      plot.subtitle = element_text(size=20, face='bold', margin=margin(0,0,10,0), hjust=0.5),
      axis.text=element_text(size=rel(1.75), color='black'),
      axis.title=element_text(size=rel(1.95), color='black'),
      axis.title.y = element_text(margin=margin(0,10,0,0)),
      axis.title.x = element_text(margin=margin(10,0,0,0)),
      legend.position="right",
      panel.border=element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank()
    )
}

# Scatter plot function
scatter_plot <- function(data, xlabel="", ylabel="", title="", subtitle="", lm=FALSE, loess = FALSE) {
  p <- ggplot(data, aes_string(x=names(data)[1], y=names(data)[2], col=names(data)[3])) +
    labs(x=xlabel, y=ylabel, title=title, subtitle=subtitle) +
    geom_point(shape = 16, size = 2.5, show.legend = FALSE, alpha = .55) + 
    scatter_theme() + 
    scale_color_gradient(low = "#0091ff", high = "#f0650e")
  if (lm){
    if (loess){ p <- p + geom_smooth() }#+ scale_x_continuous(trans = "log1p")
    else{ p <- p + geom_smooth(method='lm') }
  }
}

# Define ggplot2 boxplot theme for boxplots
boxplot_theme <- function() {
  p <- theme(
    plot.title = element_text(size=30, hjust=0.5, margin=margin(0,0,20,0)),
    axis.title.y = element_text(colour="black", size=20, vjust=1.5),
    axis.title.x = element_text(colour="black", size=20, vjust=1.5, margin=margin(15,0,0,0)),
    axis.text.x = element_text(colour="black",size=rel(1.6)),
    axis.text.y = element_text(colour="black",size=rel(1.6)),
    axis.line = element_line(colour="black", size=rel(0.7)),
    axis.ticks.x = element_line(colour="black", size=rel(0.8)),
    axis.ticks.y = element_blank(),
    panel.background = element_blank(),
    panel.grid = element_blank(),
    legend.position="none",
    legend.text=element_text(size=15),
    legend.title=element_blank(),
    legend.background=element_blank(),
    panel.border = element_blank()
  )
}

# Define ggplot2 boxplot theme for boxplots
barplot_theme <- function() {
  p <- theme(
    plot.title = element_text(size=30, hjust=0.5, margin=margin(0,0,20,0)),
    axis.title.y = element_text(colour="black", size=20, vjust=1.5),
    axis.title.x = element_text(colour="black", size=20, vjust=1.5, margin=margin(15,0,0,0)),
    axis.text.x = element_text(colour="black",size=rel(1.6)),
    axis.text.y = element_text(colour="black",size=rel(1.6)),
    axis.line = element_line(colour="black", size=rel(0.7)),
    axis.ticks.x = element_line(colour="black", size=rel(0.8)),
    axis.ticks.y = element_blank(),
    panel.background = element_blank(),
    panel.grid = element_blank(),
    legend.position="right",
    legend.text=element_text(size=13),
    legend.title=element_text(size=18),
    legend.background=element_blank(),
    panel.border = element_blank()
  )
}

# Define ggplot2 boxplot theme for boxplots
barplot_multi_theme <- function() {
  p <- theme(
    plot.title = element_text(size=30, hjust=0.5, margin=margin(0,0,20,0)),
    axis.title.y = element_text(colour="black", size=20, vjust=1.5),
    axis.title.x = element_text(colour="black", size=20, vjust=1.5, margin=margin(15,0,0,0)),
    axis.text.x = element_text(colour="black",size=rel(1.2)),
    axis.text.y = element_text(colour="black",size=rel(1.5)),
    axis.line = element_line(colour="black", size=rel(0.7)),
    axis.ticks.x = element_line(colour="black", size=rel(0.8)),
    axis.ticks.y = element_blank(),
    panel.background = element_blank(),
    panel.grid = element_blank(),
    legend.position="right",
    legend.text=element_text(size=13),
    legend.title=element_text(size=18),
    legend.background=element_blank(),
    panel.border = element_blank()
  )
}

# Define ggplot2 theme for line plots
line_theme <- function(){
  p <- theme(
      plot.title=element_text(size=25, face='bold', margin=margin(0,0,5,0), hjust=0.5),
      axis.text=element_text(size=rel(1.15), color='black'),
      axis.title=element_text(size=rel(1.55), color='black'),
      axis.title.y = element_text(margin=margin(0,15,0,0)),
      axis.title.x = element_text(margin=margin(15,0,0,0)),
      axis.ticks.x = element_line(colour="black", size=rel(0.8)),
      axis.ticks.y = element_blank(),
      legend.position="right",
      legend.key.size = unit(1.9, 'lines'),
      legend.title=element_text(size=24, face='bold'),
      legend.text=element_text(size=19),
      panel.border=element_blank(),
      panel.grid.major = element_line(colour = "gainsboro"),
      #panel.grid.minor = element_line(colour = "grey"),
      panel.background = element_blank()
    )
}
```

<!-- # Parse and filter data -->
```{r echo=FALSE, include=FALSE}
# Data
io                     <- list()
io$base_dir            <- "/home/cakapourani/Documents/Code/datasets/nmt-seq-data-eb"
# io$base_dir          <- "/Users/ricard/data/NMT-seq"
# io$base_dir          <- "/Users/kapouranis/Documents/Andreas/Code/datasets/nmt-seq-data"
io$out_dir             <- "/home/cakapourani/Documents/Code/nmt-seq/rebuttal/EB/acc_conservation/out/cluster/"
io$rna_file            <- paste0(io$base_dir, "/rna/parsed/sceset.rds")
io$acc_file_rate       <- paste0(io$base_dir, "/acc/parsed/acc_data.tsv.gz")
io$acc_file            <- paste0(io$base_dir, "/acc/parsed/profiles/cons_cluster_prom_200_200_basis13_GpCcov10_bic2_cellcov0.6.rds")
io$acc_cellwise        <- paste0(io$base_dir, "/acc/parsed/profiles/cell_wise_cl_prom_200_200_basis11_GpCcov10.rds")
io$pseudotime          <- "/home/cakapourani/Documents/Code/nmt-seq/rebuttal/EB/pseudotime/out/destiny.tsv"
io$genelist_dir        <- paste0(io$base_dir, "/metadata")
io$plur_genes_file     <- paste0(io$genelist_dir, "/pluripotency_extended.tsv")
io$h3k4me3_genes_file  <- paste0(io$genelist_dir, "/H3K4me3.bed")
io$h3k27me3_genes_file <- paste0(io$genelist_dir, "/H3K27me3.bed")

opts                   <- list()
opts$annos             <- c("prom_200_200")
opts$min.weight        <- 10      # Minimum weight per feature (met coverage)
opts$min.s             <- 10      # minimum number of samples to do the correlation
opts$gene_window       <- 5e3     # window length for the overlap between genes and features
opts$threshold         <- 5e-2    # threshold for significant hits
opts$quan_seq          <- seq(0, 1, length.out = 6) # Discretization steps
opts$axis_labels       <- paste0(seq(1,5))     # Discretization steps
```

<!-- # Load scRNA data -->
```{r load_rna, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE}
# Load expression as SCESet object
sceset        <- readRDS(file = io$rna_file)
# Create data.table
rna_dt        <- exprs(sceset) %>% t %>% as.data.table(keep.rownames = "sample") %>% melt(id.vars = "sample", value.name = "expr", variable.name = "gene")
# Compute summary statistics for each gene across cells
rna_all_stats <- rna_dt[, .(mean_expr = mean(expr), median_expr = median(expr), var_expr = var(expr), sd_expr = sd(expr), cv_expr = sd(expr)/mean(expr)), by = gene]
N_cells       <- length(unique(rna_dt$sample))  # Number of unique cells
rna_all_stats <- rna_all_stats[, cv_expr := log2((1 + 1/(4 * N_cells)) * cv_expr)]  # Correct CV empirically
# Extract biological variance, i.e. overdispersion
rna_all_stats <- rna_all_stats[, overdispersion := fData(sceset)$bioVar]
# Extract gene coordinates metadata to do the overlap
rna_metadata  <- fData(sceset) %>% tibble::rownames_to_column("gene") %>% as.data.table %>% .[,c("chr", "start", "end", "gene", "ens_id")]  %>% .[, chr := as.factor(sub("chr", "", chr))] %>% setnames("ens_id", "id")
# Merge rna data with metadata
rna_all_stats <- merge(rna_metadata[, c("chr", "id", "gene")], rna_all_stats, by = c("gene")) # Merge using all samples
rna_dt        <- merge(rna_metadata[, c("chr", "id", "gene")], rna_dt, by = c("gene"))        # Merge using only samples that pass filtering
rm(sceset, N_cells)
```

<!-- # Load acc rate data -->
```{r load_acc_rate, eval=TRUE, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE}
# Load accessibility data
acc_rate   <- fread(sprintf("zcat < %s", io$acc_file_rate)) %>% as.data.table %>% .[anno %in% opts$annos] %>% .[, rate := rate + rtruncnorm(.N, a = 0, b = Inf, mean = 0, sd = 0.01 )]
# Keep only covered regions
acc_rate   <- acc_rate[weight >= opts$min.weight] 
# Filter regions not covered across cells
keep_sites <- acc_rate[,.(n = .N), by = c("id")] %>% .[n >= opts$min.s] %>% .$id  
acc_rate   <- acc_rate[id %in% keep_sites]
# Compute summary statistics
acc_rate_stats <- acc_rate[, .(mean_acc = mean(rate), median_acc = median(rate), var_acc = var(rate), sd_acc = sd(rate)), by = id] 
rm(keep_sites)
```

<!-- # Load acc profile data -->
```{r load_acc_prof, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE}
# Load accessibility data
acc_profile  <- readRDS(io$acc_file) %>% .[cells > opts$min.s] 
# Make clusters column as factor and keep accessibility groups with more than 30 genes in each cluster
acc_profile <- acc_profile %>% .[, factor_clusters := factor(clusters)] %>% .[, N := .N, by = clusters] %>% .[N > 30]
# Merge with expression data using all samples 
accrna_prof <- merge(rna_all_stats, acc_profile, by = c("id")) %>% setorder(clusters) 
rm(acc_profile)
```

<!-- # Merge all data in one data.table object -->
```{r merge_dt, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE}
# Merge data with rate
accrna <- merge(accrna_prof, acc_rate_stats, by = c("id")) %>% setorder(clusters)
# Create accessibility variability levels using quartiles
accrna <- accrna %>% .[, acc_var_levels := cut(abs(sd_acc), breaks = quantile(abs(sd_acc), probs = opts$quan_seq), labels = opts$axis_labels, include.lowest = TRUE)]
# Create mean expression levels
accrna <- accrna %>% .[, scaled_expr := BPRMeth:::.minmax_scaling(mean_expr, fmin = 0, fmax = 1)] %>% .[, scaled_mean_expr_levels := cut(scaled_expr, seq(0 , 1, length.out = 6), labels=c(1:5), include.lowest = TRUE)]
# Create mean expression levels
accrna <- accrna %>% .[, mean_expr_levels := cut(mean_expr, breaks = quantile(mean_expr, probs = opts$quan_seq), labels = opts$axis_labels, include.lowest = TRUE)]
# Create discrete expression levels
opts$discr_labels <- c("Zero Expr", "Low Expr", "Medium Expr", "High Expr")     # Discretization steps
accrna <- accrna %>% .[, discr_expr_levels := cut(mean_expr, breaks = c(0 , 0.001, 2, 6, 20), labels = opts$discr_labels, include.lowest = TRUE)]
```

# Accessibility conservation and expression relationship

The region for the accesibility profiles is $\pm 200$ around TSS. 

## Association with mean expression
We observe that accessibility conservation around promoter regions is positively correlated with mean expression levels.
```{r associate, echo=FALSE, warning=FALSE, message=FALSE}
# Acc profile conservation - Mean expression association
p <- ggplot(accrna[,c("factor_clusters", "mean_expr", "N")], aes(x=factor_clusters, y=mean_expr)) +
  geom_boxplot(alpha=0.7, outlier.shape=NA, fill="#00BFC4") + geom_hline(yintercept=0, colour="black", linetype="dashed") +
  stat_summary(fun.y = function(x){ xx <- min(x); return(xx - 0.7)}, aes(label = paste0("N=", as.character(N))), geom='text', col='black', size = 5) +
  labs(title=NULL, x="Clusters (i.e. conservation)", y="Mean expression") + boxplot_theme()
p
invisible(pdf(file = paste0(io$out_dir, "box_accProfCons_exprMean.pdf"), width = 9, height = 6, useDingbats=FALSE)); p; invisible(dev.off())

# Acc rate variance - Mean expression association
p <- ggplot(accrna[,c("acc_var_levels", "mean_expr")], aes(x=acc_var_levels, y=mean_expr)) +
  geom_boxplot(alpha=0.7, outlier.shape=NA, fill="#00BFC4") + geom_hline(yintercept=0, colour="black", linetype="dashed") +
  labs(title=NULL, x="Accessibility rate variance", y="Mean expression") + boxplot_theme()
p
invisible(pdf(file = paste0(io$out_dir, "box_accRateVar_exprMean.pdf"), width = 9, height = 6, useDingbats=FALSE)); p; invisible(dev.off())
rm(p)
```

Here we show the relationship between the accessibility conservation profiles and mean accessibility rate.
```{r associate_2, echo=FALSE, warning=FALSE, message=FALSE,fig.width=9, fig.height=9}
# Acc profile conservation - Acc rate association
p1 <- ggplot(accrna[,c("factor_clusters", "mean_acc", "N")], aes(x=factor_clusters, y=mean_acc)) +
  geom_boxplot(alpha=0.7, outlier.shape=NA, fill="#00BFC4") + 
  stat_summary(fun.y = function(x){ xx <- 0; return(xx + 0.7)}, aes(label = paste0("N=", as.character(N))), geom='text', col='black', size = 5) +
  labs(title=NULL, x="Clusters (i.e. conservation)", y="Accessibility rate") + boxplot_theme()
# Acc profile conservation - Acc rate association
p2 <- ggplot(accrna[,c("factor_clusters", "var_acc", "N")], aes(x=factor_clusters, y=var_acc)) +
  geom_boxplot(alpha=0.7, outlier.shape=NA, fill="#00BFC4") + 
  stat_summary(fun.y = function(x){ xx <- 0; return(xx + 0.7)}, aes(label = paste0("N=", as.character(N))), geom='text', col='black', size = 5) +
  labs(title=NULL, x="Clusters (i.e. conservation)", y="Acc rate variance") + boxplot_theme()
# Combine them to create one plot
pp <- cowplot::plot_grid(p1, p2, label_size = 30, labels="AUTO", ncol = 1, nrow = 2, rel_widths = c(1, 1))
plot(pp)
invisible(pdf(file = paste0(io$out_dir, "box_accProfCons_accRate.pdf"), width = 9, height = 9, useDingbats=FALSE)); pp; invisible(dev.off())
rm(p1, p2, pp)
```

Below we show the relationship between mean accessibility rate and mean expression levels across cells, and as expected we have a positive association between these layers.
```{r associate_3, echo=FALSE, warning=FALSE, message=FALSE}
# Acc rate - Mean expression association
p <- scatter_plot(accrna[, c("mean_expr", "mean_acc", "var_acc")], xlabel = "Mean expression", ylabel = "Accessibility rate", title = "", subtitle = paste0("Pearson's r = ", round(cor(accrna$mean_acc, accrna$mean_expr), 2)), lm = TRUE)
p
invisible(pdf(file = paste0(io$out_dir, "scatter_accRate_exprMean.pdf"), width = 9, height = 6, useDingbats=FALSE)); p; invisible(dev.off())

p <- scatter_plot(acc_rate[, c("rate", "weight", "var")], xlabel = "Accessibility rate", ylabel = "Coverage", title = "", subtitle = paste0("Pearson's r = ", round(cor(acc_rate$rate, acc_rate$weight), 2)), lm = TRUE)
p
invisible(pdf(file = paste0(io$out_dir, "scatter_accRate_accCoverage_sc.pdf"), width = 9, height = 6, useDingbats=FALSE)); p; invisible(dev.off())

rm(p)
```

## Association with expression variability

```{r mean_var_expr, eval=FALSE, echo=FALSE, warning=FALSE, message=FALSE, fig.width=9, fig.height=12}
# Expression mean - variance relationship
p1 <- scatter_plot(accrna[, c("mean_expr", "var_expr", "clusters")], xlabel = "Mean expression", ylabel = "Expr variance", title = "", lm = TRUE, loess = TRUE)
# Expression mean - CV relationship
p2 <- scatter_plot(accrna[, c("mean_expr", "cv_expr", "clusters")], xlabel = "Mean expression", ylabel = "Expr log2CV", title = "", lm = TRUE, loess = TRUE)
# Expression mean - overdispersion relationship
p3 <- scatter_plot(accrna[, c("mean_expr", "overdispersion", "clusters")], xlabel = "Mean expression", ylabel = "Expr overdisp", title = "", lm = TRUE, loess = TRUE)
pp <- cowplot::plot_grid(p1, p2, p3, label_size = 30, labels="AUTO", ncol = 1, nrow = 3, rel_widths = c(1, 1))
plot(pp)
invisible(pdf(file = paste0(io$out_dir, "scatter_exprStats.pdf"), width = 9, height = 12, useDingbats=FALSE)); print(pp); invisible(dev.off())
rm(p1, p2, p3, pp)
```

# GO analysis
Here we perform Gene Ontology analysis to see whether specific clusters of genes are enriched for specific pathways.
```{r go_enr_list, echo=FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=6}
io$reactomeFile  <- paste0(io$genelist_dir, "/mouse_v75_reactome.rds")
io$reactomeFile2 <- paste0(io$genelist_dir, "/AllPathways.txt")

# Read reactome file and change gene names to Upper case
reactome           <- readRDS(io$reactomeFile); colnames(reactome) <- toupper(colnames(reactome))
# Read pathways file to change from ID to description pathway
tmp                <- read.table(io$reactomeFile2, header=F, quote="", sep="\t", stringsAsFactors=F)[,c(1,2)]
reactome_meta      <- tmp[,2]; names(reactome_meta) <- tmp[,1]
rownames(reactome) <- stringr::str_replace_all(rownames(reactome), reactome_meta)
feature.sets       <- reactome  # Store results to the following object
opts$min.size      <- 10  # Filter feature sets with small number of features
feature.sets       <- feature.sets[rowSums(feature.sets)>=opts$min.size, ]
# Remove features that do not intersect between the feature sets and the data
accrna_tmp        <- copy(accrna)
features          <- intersect(toupper(accrna_tmp$gene), colnames(feature.sets))
foo_filt          <- accrna_tmp[,gene := toupper(gene)] %>% .[gene %in% features]
feature.sets_filt <- feature.sets[,features]

# Melt and merge data to do the fisher's test
asd <- melt.data.table(as.data.table(feature.sets_filt, keep.rownames="pathway"), id.vars = "pathway", 
                       variable.name="gene", value.name="value") %>% merge(foo_filt[,c("gene","clusters")], allow.cartesian=T, by="gene")
# Get GO list for each cluster
go_list <- list()
for (c in unique(sort(asd$cluster))) {
  go_list[[c]] <- asd[,.(p=fisher.test(x=matrix(data=c(.SD[clusters==c & value==1,.N], .SD[clusters!=c & value==1,.N],
                                                       .SD[clusters==c & value==0,.N], .SD[clusters!=c & value==0,.N]), 
                                                nrow=2, ncol=2), alternative="greater")[["p.value"]]), by="pathway"] %>% 
    .[,padj:=p.adjust(p, method = "fdr")] %>% setorder(p)
  # Create a nice formatted table
  # print(knitr::kable(go_list[[c]][1:45], caption = paste0("GO enrichment: Cluster ", c)))
  
  # Convert pvalues to log scale (add a small regulariser to avoid numerical errors)
  tmp <- copy(go_list[[c]])
  tmp$log <- -log10(tmp$padj)
  #order according to significance
  tmp$pathway <- factor(tmp$pathway, levels = tmp$pathway[order(tmp$padj, decreasing = T)])
  p <- ggplot(tmp[1:38], aes(x=pathway, y=log)) +
    geom_point(size=5) +
    geom_hline(yintercept=-log10(opts$threshold), linetype="longdash", color = "red") +
    scale_color_manual(values=c("black","red")) +
    geom_segment(aes(xend=pathway, yend=0)) +
    ylab(expression(paste("-log"[10],"(",plain(q),"-value)"))) + coord_flip() +
    theme(axis.text.y = element_text(size=rel(0.9), hjust=1, color='black'),
      axis.text.x = element_text(size=rel(1.2), vjust=0.5, color='black'),
      axis.title.y=element_blank(), legend.position='none',
      panel.background = element_blank()
    )
  print(p)
  invisible(pdf(file = paste0(io$out_dir, "GO_analysis/GO_enrichment_C", c, ".pdf"), width = 12, height = 10, useDingbats=FALSE)); print(p); invisible(dev.off())
  fwrite(go_list[[c]], file = paste0(io$out_dir, "GO_analysis/GO_enrichment_C", c, ".csv"))
}
saveRDS(go_list, file = paste0(io$out_dir, "GO_analysis/GO_list.rds"))
rm(reactome, reactome_meta, tmp, p, asd, features, foo_filt, feature.sets_filt, accrna_tmp, feature.sets, c)
```

## GO analysis using variance of accessibility rate
Here we perform Gene Ontology analysis to see whether specific clusters of genes are enriched for specific pathways.
```{r go_enr_list_rate, echo=FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=6}
io$reactomeFile  <- paste0(io$genelist_dir, "/mouse_v75_reactome.rds")
io$reactomeFile2 <- paste0(io$genelist_dir, "/AllPathways.txt")

# Read reactome file and change gene names to Upper case
reactome           <- readRDS(io$reactomeFile); colnames(reactome) <- toupper(colnames(reactome))
# Read pathways file to change from ID to description pathway
tmp                <- read.table(io$reactomeFile2, header=F, quote="", sep="\t", stringsAsFactors=F)[,c(1,2)]
reactome_meta      <- tmp[,2]; names(reactome_meta) <- tmp[,1]
rownames(reactome) <- stringr::str_replace_all(rownames(reactome), reactome_meta)
feature.sets       <- reactome  # Store results to the following object
opts$min.size      <- 10  # Filter feature sets with small number of features
feature.sets       <- feature.sets[rowSums(feature.sets)>=opts$min.size, ]
# Remove features that do not intersect between the feature sets and the data
accrna_tmp        <- copy(accrna)
features          <- intersect(toupper(accrna_tmp$gene), colnames(feature.sets))
foo_filt          <- accrna_tmp[,gene := toupper(gene)] %>% .[gene %in% features]
feature.sets_filt <- feature.sets[,features]

# Melt and merge data to do the fisher's test
asd <- melt.data.table(as.data.table(feature.sets_filt, keep.rownames="pathway"), id.vars = "pathway", 
                       variable.name="gene", value.name="value") %>% merge(foo_filt[,c("gene","acc_var_levels")], allow.cartesian=T, by="gene")
# Get GO list for each cluster
go_list <- list()
for (c in unique(sort(asd$acc_var_levels))) {
  go_list[[c]] <- asd[,.(p=fisher.test(x=matrix(data=c(.SD[acc_var_levels==c & value==1,.N], .SD[acc_var_levels!=c & value==1,.N],
                                                       .SD[acc_var_levels==c & value==0,.N], .SD[acc_var_levels!=c & value==0,.N]), 
                                                nrow=2, ncol=2), alternative="greater")[["p.value"]]), by="pathway"] %>% 
    .[,padj:=p.adjust(p, method = "fdr")] %>% setorder(p)
  # Create a nice formatted table
  # print(knitr::kable(go_list[[c]][1:45], caption = paste0("GO enrichment: Cluster ", c)))
  
  # Convert pvalues to log scale (add a small regulariser to avoid numerical errors)
  tmp <- copy(go_list[[c]])
  tmp$log <- -log10(tmp$padj)
  #order according to significance
  tmp$pathway <- factor(tmp$pathway, levels = tmp$pathway[order(tmp$padj, decreasing = T)])
  p <- ggplot(tmp[1:38], aes(x=pathway, y=log)) +
    geom_point(size=5) +
    geom_hline(yintercept=-log10(opts$threshold), linetype="longdash", color = "red") +
    scale_color_manual(values=c("black","red")) +
    geom_segment(aes(xend=pathway, yend=0)) +
    ylab(expression(paste("-log"[10],"(",plain(q),"-value)"))) + coord_flip() +
    theme(axis.text.y = element_text(size=rel(0.9), hjust=1, color='black'),
      axis.text.x = element_text(size=rel(1.2), vjust=0.5, color='black'),
      axis.title.y=element_blank(), legend.position='none',
      panel.background = element_blank()
    )
  print(p)
  invisible(pdf(file = paste0(io$out_dir, "GO_analysis/rate/GO_enrichment_C", c, ".pdf"), width = 12, height = 10, useDingbats=FALSE)); print(p); invisible(dev.off())
  fwrite(go_list[[c]], file = paste0(io$out_dir, "GO_analysis/rate/GO_enrichment_C", c, ".csv"))
}
saveRDS(go_list, file = paste0(io$out_dir, "GO_analysis/rate/GO_list.rds"))
#rm(reactome, reactome_meta, tmp, p, asd, features, foo_filt, feature.sets_filt, accrna_tmp, feature.sets, c)
```


# Histone marks
Here we perform Histone mark enrichment analysis to see whether specific clusters of genes are enriched for specific marks.
```{r hm_genelist, echo=FALSE, warning=FALSE, message=FALSE, results='asis',fig.width=12, fig.height=7}
# Read histone marks files
h3k4_genes <- fread(io$h3k4me3_genes_file) %>% as.data.table %>% .[, c("V1", "V2", "V3")] %>% setnames(c("chr", "start", "end")) %>% setkey(chr, start, end) %>% .[, chr := as.factor(sub("chr", "", chr))] %>% .[, id := paste0("hek4m3_", 1:.N)]
h3k27_genes <- fread(io$h3k27me3_genes_file) %>% as.data.table %>% .[, c("V1", "V2", "V3")] %>% setnames(c("chr", "start", "end")) %>% setkey(chr, start, end) %>% .[, chr := as.factor(sub("chr", "", chr))] %>% .[, id := paste0("h3k27me3_", 1:.N)]
# Get RNA metdata to do the overlapping with genes
rna_metadata_tmp <- rna_metadata[, c("chr", "start", "end", "gene")] %>%
  .[,c("start", "end") := list(start - opts$gene_window, start + opts$gene_window)] %>% setkey(chr, start, end)
# Do the overlap
ov_h3k4  <- foverlaps(h3k4_genes, rna_metadata_tmp, nomatch = 0) %>% .[,c("gene")] %>% unique
ov_h3k27 <- foverlaps(h3k27_genes, rna_metadata_tmp, nomatch = 0) %>% .[,c("gene")] %>% unique
ov_h3    <- Reduce(intersect, list(ov_h3k4$gene, ov_h3k27$gene)) %>% as.data.table %>% setnames(c("gene"))
# Total number of clusters k
cls      <- length(unique(accrna$clusters))
h3k4_c = h3k27_c = biv_c = unmark_c <- vector(mode = "numeric", length = cls)
for (i in 1:cls){  # Iterate over each cluster and extract number of genes related to each histone mark
  h3k4_c[i]   <- NROW(accrna[gene %in% ov_h3k4$gene & !(gene %in% ov_h3k27$gene) & clusters == i])
  h3k27_c[i]  <- NROW(accrna[gene %in% ov_h3k27$gene & !(gene %in% ov_h3k4$gene) & clusters == i])
  biv_c[i]    <- NROW(accrna[gene %in% ov_h3$gene & clusters == i])
  unmark_c[i] <- NROW(accrna[!(gene %in% ov_h3k27$gene | gene %in% ov_h3k4$gene) & clusters == i])
}

# Create object for plotting
dt <- data.table(Cluster = rep(paste0("C", seq(1:cls)), 4), 
                 Mark = c(rep("H3K4me3", cls), rep("H3K27me3", cls), rep("Bivalent", cls), rep("Unmarked", cls)), 
                 Proportion = c(as.integer(h3k4_c), as.integer(h3k27_c), as.integer(biv_c), as.integer(unmark_c)))
dt <- dt %>% .[, N := sum(Proportion), by = Cluster]

p <- ggplot(dt, aes(fill=Mark, y=Proportion, x=Cluster)) + geom_bar(stat="identity", position="fill", alpha=1) + 
  stat_summary(fun.y = function(x){ xx <- 0; return(xx - 0.05)}, aes(label = paste0("N=", as.character(N))), geom='text', col='black', size = 7) +
  #scale_fill_manual(values = c("tomato", "yellow2", "chartreuse3", "cornflowerblue")) 
  labs(x="Clusters", y="Relative Proportion", fill = "Histone mark") +
  scale_fill_discrete(name="Histone mark") + 
  scale_fill_brewer(palette = "Set3") + barplot_theme()
p
invisible(pdf(file = paste0(io$out_dir, "HM_analysis/bar_HM_enrichment.pdf"), width = 12, height = 7)); p; invisible(dev.off())
rm(h3k4_genes, h3k27_genes, rna_metadata_tmp, p, i, unmark_c, h3k4_c, h3k27_c, biv_c, cls)
```


## Histone marks by mean expression levels discretized
```{r hm_genelist_discr, echo=FALSE, warning=FALSE, message=FALSE, results='asis',fig.width=12, fig.height=6}
# Total number of clusters k
cls      <- length(unique(accrna$clusters))
disc_lev <- levels(accrna$discr_expr_levels)
h3k4_c = h3k27_c = biv_c = unmark_c <- vector(mode = "numeric", length = cls)
dt_all <- data.table(Cluster = character(), 
                 Mark = character(), 
                 Proportion = numeric(),
                 Mean_Level = factor())
# m <- matrix(0, nrow = cls * length(disc_lev), ncol = 4)
# kk <- 1
for (i in 1:cls){  # Iterate over each cluster and extract number of genes related to each histone mark
  for (l in disc_lev){
    h3k4_c[i]   <- NROW(accrna[gene %in% ov_h3k4$gene & !(gene %in% ov_h3k27$gene) & clusters == i & discr_expr_levels == l])
    h3k27_c[i]  <- NROW(accrna[gene %in% ov_h3k27$gene & !(gene %in% ov_h3k4$gene) & clusters == i & discr_expr_levels == l])
    biv_c[i]    <- NROW(accrna[gene %in% ov_h3$gene & clusters == i & discr_expr_levels == l])
    unmark_c[i] <- NROW(accrna[!(gene %in% ov_h3k27$gene | gene %in% ov_h3k4$gene) & clusters == i & discr_expr_levels == l])
    
    # print(c(h3k4_c[i], h3k27_c[i], biv_c[i], unmark_c[i]))
    # m[kk, ] <- c(h3k4_c[i], h3k27_c[i], biv_c[i], unmark_c[i])
    # kk <- kk + 1
    # Create object for plotting
    dt <- data.table(Cluster = paste0("C", i), 
                   Mark = c("H3K4me3", "H3K27me3", "Bivalent", "Unmarked"), 
                   Proportion = c(as.integer(h3k4_c[i]), as.integer(h3k27_c[i]), as.integer(biv_c[i]), as.integer(unmark_c[i])),
                   Mean_Level = l)
    dt_all <- rbind(dt_all, dt)
  }
}
##colnames(m) <- c("H3K4", "H3K27", "Bivalent", "Unmarked")
dt_all <- dt_all %>% .[, N := sum(Proportion), by = c("Cluster", "Mean_Level")]


p <- ggplot(dt_all, aes(fill=Mark, y=Proportion, x=Cluster)) + geom_bar(stat="identity", position="fill", alpha=1) + 
  facet_grid(. ~ Mean_Level) +
  stat_summary(fun.y = function(x){ xx <- 0; return(xx - 0.05)}, aes(label = paste0("N=", as.character(N))), geom='text', col='black', size = 3.7) +
  labs(x="Clusters", y="Relative Proportion", fill = "Histone mark") +
  scale_fill_discrete(name="Histone mark") + 
  scale_fill_brewer(palette = "Set3") + barplot_multi_theme() + theme(legend.position="none")
p
invisible(pdf(file = paste0(io$out_dir, "HM_analysis/bar_HM_enrichment_discrete.pdf"), width = 12, height = 6)); p; invisible(dev.off())
```


## Histone marks by mean expression levels quantile
```{r hm_genelist_quant, echo=FALSE, warning=FALSE, message=FALSE, results='asis',fig.width=13, fig.height=5.5}
# Total number of clusters k
cls      <- length(unique(accrna$clusters))
mean_lev <- length(unique(accrna$mean_expr_levels))
h3k4_c = h3k27_c = biv_c = unmark_c <- vector(mode = "numeric", length = cls)
dt_all <- data.table(Cluster = character(), 
                 Mark = character(), 
                 Proportion = numeric(),
                 Mean_Level = factor())
for (i in 1:cls){  # Iterate over each cluster and extract number of genes related to each histone mark
  for (l in 1:mean_lev){
    h3k4_c[i]   <- NROW(accrna[gene %in% ov_h3k4$gene & !(gene %in% ov_h3k27$gene) & clusters == i & mean_expr_levels == l])
    h3k27_c[i]  <- NROW(accrna[gene %in% ov_h3k27$gene & !(gene %in% ov_h3k4$gene) & clusters == i & mean_expr_levels == l])
    biv_c[i]    <- NROW(accrna[gene %in% ov_h3$gene & clusters == i & mean_expr_levels == l])
    unmark_c[i] <- NROW(accrna[!(gene %in% ov_h3k27$gene | gene %in% ov_h3k4$gene) & clusters == i & mean_expr_levels == l])
    
    # print(c(h3k4_c[i], h3k27_c[i], biv_c[i], unmark_c[i]))
    # Create object for plotting
    dt <- data.table(Cluster = paste0("C", i), 
                   Mark = c("H3K4me3", "H3K27me3", "Bivalent", "Unmarked"), 
                   Proportion = c(as.integer(h3k4_c[i]), as.integer(h3k27_c[i]), as.integer(biv_c[i]), as.integer(unmark_c[i])),
                   Mean_Level = paste0("Expr Q", l))
    dt_all <- rbind(dt_all, dt)
  }
}
dt_all <- dt_all %>% .[, N := sum(Proportion), by = c("Cluster", "Mean_Level")]

p <- ggplot(dt_all, aes(fill=Mark, y=Proportion, x=Cluster)) + geom_bar(stat="identity", position="fill", alpha=1) + 
  facet_grid(. ~ Mean_Level) +
  stat_summary(fun.y = function(x){ xx <- 0; return(xx - 0.05)}, aes(label = paste0("N=", as.character(N))), geom='text', col='black', size = 3.3) +
  labs(x="Clusters", y="Relative Proportion", fill = "Histone mark") +
  scale_fill_discrete(name="Histone mark") + 
  scale_fill_brewer(palette = "Set3") + barplot_multi_theme() + theme(legend.position="none")
p
invisible(pdf(file = paste0(io$out_dir, "HM_analysis/bar_HM_enrichment_quantile.pdf"), width = 13, height = 5.5)); p; invisible(dev.off())
```

# Pluripotency genes
Here we want to observe how pluripotent genes map to each cluster, i.e. accessibility conservation.
```{r plur_genelist, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
# Read pluripotency genes file
plur_genes <- fread(io$plur_genes_file) %>% as.data.table %>% setnames(c("id", "gene"))
plur_genes <- merge(plur_genes, accrna, by = c("id", "gene"))
print(plur_genes[, .N, by = factor_clusters])
```

